---
title: "Multivariate Analysis Task 1"
author: "Nzambuli Daniel"
date: "2024-01-27"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Question

Air pollution is currently one of the most serious public health worries
worldwide. Many epidemiological studies have proved the influence that
some chemical compounds can have on our health. Associations set up to
monitor air quality are active all over the world to measure the
concentration of these pollutants. They also keep a record of
meteorological conditions such as temperature (T9, T12, T15), cloud
cover (C9, C12, C15), wind (W9, W12, W15). The data set is within the
`fdm2id` package.

The final aim is to analyse the relationship between the maximum daily
ozone level and meteorological conditions. You have at our disposal 112
observations collected. You need to inspect the data with a PCA and
illustrate the reconstruction property of PCA.

The variables available are

-   max O3 (maximum daily ozone),

-   Temperature at 0900HRS, 1200HRS & 1500HRS (T9, T12, T15)

-   cloud cover at 0900HRS, 1200HRS & 1500HRS (C9, C12, C15)

-   wind speed at 0900HRS, 1200HRS & 1500HRS (W9, W12, W15)

### The package

```{r}
library(fdm2id)
```

### The data 

```{r}
data(airquality)
head(airquality)
```

## Perform the PCA on the ozone data and give a short interpretation of the dimensions of variability (interpretation of the observations plot, correlation between variables, percentage of variability, etc.

### EDA

```{r}
any(is.na(airquality))
```

```{r}
data = airquality
# quantity of data
cat("number of rows:", nrow(data), "\nnumber of cols:", ncol(data), "\nnumber of values:", nrow(data) * ncol(data))
```

The required data is 112 rows. Trying to omit `na` values to see if it
fits

```{r}
data = na.omit(airquality)
nrow(data)
```

There is a missing row.

### Correlation matrix

```{r}
cor_mat =cor(data)  
cor_mat
```

### Plot for clear view

```{r}
library(corrplot)

corrplot(cor_mat)
```

> **From this**
>
> -   temperature and ozone have a `high positive` correlation
>
> -   temperature and wind have a `high negative` correlation
>
> -   wind and ozone have a very `highly negative` correlation. As wind
>     increases ozone degrading reduces. This can be a result of the
>     wind lowering the temperature.
>
> -   solar radiation has a `small positive` correlation to ozone
>     degrading and an equally sized `positive relationship` to
>     temperature
>
> Seeing how the combined effect of temperature and ozone degrading
> impacts solar radiation will be interesting.

## PCA analysis

### Principle components(pc)

expecting `6` pc

```{r}
# Get the pc
princ_comp = prcomp(data, scale = TRUE) #mean 0 and sd 1

# extract the eigen vectors
princ_comp$rotation
```

### Correct eigen value direction

```{r}
eig_val = -1 * princ_comp$rotation
eig_val
```

**Initial Overview**

> **PC1**
>
> -   made up of mostly `ozone, wind, temperature`
>
> -    temperature and ozone have a `direct relationship`
>
> -   wind and ozone have an `inverse relationship`
>
> **PC2**
>
> -   made up of mostly `solar radiation and month`
>
> -   solar radiation and month have a `inverse relationship`
>
> **PC3**
>
> -   mostly made up of `day`
>
> **PC4**
>
> -   mostly `solar radiation, wind, month`
>
> **PC5**
>
> -   mostly `wind and temperature`
>
> **PC6**
>
> -   ozone and temperature

### Variance explained by each PC

```{r}
var_exp = princ_comp$sdev ^2 / sum(princ_comp$sdev ^2)
var_exp
```

### Variance Explained Table

```{r}
var_exp_tab = data.frame(PC = c("PC1",
                                "PC2",
                                "PC3",
                                "PC4",
                                "PC5",
                                "PC6"),
                         var_exp = var_exp)

# cummulative variation explained
var_exp_tab$cum_sum = cumsum(var_exp)


# percentage variation explained
var_exp_tab$perc_exp = var_exp_tab$cum_sum *100

# view 
var_exp_tab
```

### Find the Ideal PC to choose

```{r}
library(ggplot2)
qplot(seq(1:6), var_exp)+
  geom_line() + 
  labs(x ="Principal Component", 
  y="Variance Explained",
  title = "Scree Plot") +
  ylim(0, 1)
```

The key pc to select are the first 3 pc

### Selecting only key components

```{r}
eig_val[, 1:3]
```

### Plot Relationship

```{r}
library(factoextra)
fviz_pca_biplot(princ_comp,
                label = "var",
                col.ind = "black",
                col.var = "contrib",
                gradient.cols = c("blue","skyblue","coral","red"))
```

### explaining inference

> **Temperature and Time**
>
> temperature `increases` as the months progress from January
>
> As the **day** progresses the temperature `increases` . This leads to
> `higher wind speeds` that cause cooling which lowers the temperature.
>
> the **Wind** is lowest at high temperatures and very low temperatures

> **Temperature and Ozone**
>
> -   An `increase in temperature` causes a directly propotional
>     `increase` in the `degradation of the ozone` . From **PC1** the
>     ratio of temperature increase to ozone degrading is $\approx 1:1$
>     . calculated as
>
> $$
> \frac{ozone}{temp} = \frac{-0.55674991}{-0.55816827} = 0.9974589 \approx 1
> $$
>
> -   `increase` ozone degrading causes a ***extreme*** `increase` in
>     solar radiation with the most severe case being in **PC3** where a
>     `1% increase` in ozone degrading leads to a `233.1416%` increase
>     in solar radiation
